INGRID: Creating Languages in MPS from ANTLR Grammars

### 2016-11-01 chapters/language_definition.tex
Fix a few typos

The first phase in the process of MPS language construction is parsing of the input ANTLR v4 grammar.
It is done using an ANTLR parser that was automatically generated from the grammar of the ANTLR v4 notation itself.
Nevertheless, the full AST, which comes out of the parser, is quite complex and contains information not relevant for the INGRID algorithm.
In order to get a simple representation that is easy process by the later phases and keeps only information necessary for the construction of MPS languages, several steps of post-processing of the AST have to be performed in this phase.
An output of this phase is a custom AST-like structure representing the grammar.

@@

Our INGRID method targets mostly text-based languages, for whose concepts the visual representation in a projectional editor must be almost equivalent to their plain-text representation.
An obvious choice would therefore be to use the same approach for Editor and TextGen.
Nevertheless, since the Editor aspect has to be adjusted manually, we decided to use a different approach for TextGen.
We designed a procedure based on a simple fully automated heuristic that provides surprisingly good results.

The procedure creates the TextGen aspect for a given language concept (AST node) in two steps.
First, it generates a basic variant that inserts spaces in between every two tokens of the textual representation of the concept.
In the second step, which is the core of our heuristic, spaces are eliminated from places where they are not really needed.
The main criterion is whether the generated plain-text output can be safely processed by a parser of the original language (using the ANTLR grammar).
We discuss all the cases here:
\begin{itemize}
	\item When there is a non-alphabetical literal that is used as a token in the grammar, and that might get recognized by the parser without the need for whitespace separators around it, then we can omit the spaces. An example of such literal is '\textless' in SimpleXML.
	\item In the case of an arbitrary string, the whitespace may be omitted when the adjacent literal ends, respectively begins, with a non-alphabetical character.
		This applies especially to the values of properties defined in Structure aspects, such as the name of an XML element.
		Based on this heuristic, redundant spaces inside of quotes will be eliminated, as well as spaces next to semicolons and around brackets.
		On the other hand, it will distinguish language keywords (e.g., \code{function}, \code{var}, \code{in}) from other literals by preserving the space character in between them.
	\item Spaces can be safely omitted also when specific optional child nodes are not present, and in the case of empty lists of child nodes, so that whitespace does not accumulate.
\end{itemize}
A space must be alway inserted when two child nodes are next to each other.
Elements of a sequence of child nodes are separated with space or a line break.

>>>>

The first phase in the process of MPS language construction is parsing of the input ANTLR v4 grammar.
It is done using an ANTLR parser that was automatically generated from the grammar of the ANTLR v4 notation itself.
Nevertheless, the full AST, which comes out of the parser, is quite complex and contains information not relevant for the INGRID algorithm.
In order to get a simple representation that is easy to process by the later phases and keeps only information necessary for the construction of MPS languages, several steps of post-processing of the AST have to be performed in this phase.
An output of this phase is a custom AST-like structure representing the grammar.

@@

Our INGRID method targets mostly text-based languages, for the concepts of which the visual representation in a projectional editor must be almost equivalent to their plain-text representation.
An obvious choice would therefore be to use the same approach for Editor and TextGen.
Nevertheless, since the Editor aspect has to be adjusted manually, we decided to use a different approach for TextGen.
We designed a procedure based on a simple fully automated heuristic that provides surprisingly good results.

The procedure creates the TextGen aspect for a given language concept (AST node) in two steps.
First, it generates a basic variant that inserts spaces in between every two tokens of the textual representation of the concept.
In the second step, which is the core of our heuristic, spaces are eliminated from places where they are not really needed.
The main criterion is whether the generated plain-text output can be safely processed by a parser of the original language (using the ANTLR grammar).
We discuss all the cases here:
\begin{itemize}
	\item When there is a non-alphabetical literal that is used as a token in the grammar, and that might get recognized by the parser without the need for whitespace separators around it, then we can omit the spaces. An example of such literal is '\textless' in SimpleXML.
	\item In the case of an arbitrary string, the whitespace may be omitted when the adjacent literal ends, respectively begins, with a non-alphabetical character.
		This applies especially to the values of properties defined in Structure aspects, such as the name of an XML element.
		Based on this heuristic, redundant spaces inside of quotes will be eliminated, as well as spaces next to semicolons and around brackets.
		On the other hand, it will distinguish language keywords (e.g., \code{function}, \code{var}, \code{in}) from other literals by preserving the space character in between them.
	\item Spaces can be safely omitted also when specific optional child nodes are not present, and in the case of empty lists of child nodes, so that whitespace does not accumulate.
\end{itemize}
A space must be always inserted when two child nodes are next to each other.
Elements from a sequence of child nodes are separated with spaces or line breaks.
